% !TeX root = ../main.tex
\begin{englishabstract}
% Uncomment if funding information is available.
% \footnotetext{*The work was supported by the Foundation (foundation ID).}
% \astfootnote{The work was supported by the Foundation (foundation ID).}

With the widespread deployment of deep neural networks and large language models in the cloud and at the edge, demand for hardware compute is growing rapidly. To increase compute capability, neural network chips are proposed as domain-specific architectures and are evolving from single-core to multi-core designs. As chip scale grows, system-level issues such as on-chip network congestion, insufficient main-memory bandwidth, and high cross-core scheduling complexity are becoming increasingly salient. Against a backdrop of high tape-out cost and long iteration cycles, a cycle-level simulator that can provide quantitative feedback during the design phase plays an important supporting role in architecture evaluation and software--hardware co-optimization.

This thesis designs and implements an end-to-end cycle-level simulator that takes as input an intermediate representation describing neural network workloads. Through general-purpose modeling of neural network chips and a unified task abstraction, multi-core execution is cast as a parsable model of ``core-level workload---explicit dependency---data flow'', and compute, on-chip interconnect, and off-chip memory are advanced in a coordinated manner on a unified timeline. During simulation, per-core cycles in loading, computing, writeback, and network backpressure can be recorded to produce reproducible bottleneck attribution statistics, forming a closed loop of ``parse---execute---statistics''.

For validation and application, this work constructs a series of micro-benchmark experiments to verify inter-core data-dependency propagation, correct request--response behavior for the on-chip network and off-chip memory, and differences among subsystem backends. Compared with results from the Synopsys ZeBu hardware emulation platform under the same scheduling outcome and hardware configuration for workloads including ResNet34 and ResNet50, the relative error of this simulator is kept within $\pm$5.2\%, validating its accuracy. On this basis, this thesis carries out systematic bottleneck analysis and design-space exploration for multi-core neural network chips. In a four-core chip running ResNet50, progressive optimization---first upgrading off-chip memory, then increasing SRAM bandwidth---reduces end-to-end total cycles by 73.7\% (a 3.80$\times$ speedup) and reveals a stepwise bottleneck migration from off-chip bandwidth to on-chip port bandwidth. A further design-space exploration along three dimensions---single-core compute capability, core-count scaling, and batch size---with 12 configurations quantitatively shows that: (1) increasing compute capability yields no contribution to throughput when memory bandwidth is below what compute intensity demands; (2) scaling core count beyond the workload's parallelism limit severely degrades performance or even causes deadlock; and (3) batch-scaling efficiency is only about 24\%, limited by memory-bandwidth contention.

This project provides a reproducible cycle-level simulation basis for evaluating multi-core neural network chip architectures and for software--hardware co-optimization. Through a complete experimental chain spanning accuracy validation, bottleneck attribution, and multi-dimensional design-space exploration, the simulator demonstrates practical value in correctness verification, bottleneck localization, and architectural parameter tuning, and can offer reproducible evaluation capability for software--hardware co-optimization.

\englishKeywords{Neural network chip; Multi-core architecture; Cycle-level simulator; Software--hardware co-optimization}

% 每个关键词组的第一个字母大写，其余为小写，每一关键词之间用分号分开，最后一个关键词后不打标点符号。例如：Drip irrigation emitter; RP\&M; Hydraulics; Labyrinth flow channel\newline

\englishType{Applied Research}

% 须与中文摘要中的论文类型一致；每个单词第一个字母大写，其余为小写。例如：Applied Research

% 论文类型包括：a.理论研究(Theoretical Research)；b.应用基础(Application Fundamentals)；c.应用研究(Application Research)；d.研究报告(Research Report)；e.设计报告(Design Report)；f.案例分析(Case Study)；g.调研报告(Investigation Report)；h.产品研发(Product Development)；i.工程设计(Engineering Design)；j.工程/项目管理(Engineering/Project Management)；k.其它（Others）。

\end{englishabstract}
